One-shot learning: From domain knowledge to action models

### 2019-02-24 doc/ijcai-19-deductivelearning/deductivelearning.tex
New updates

\begin{enumerate}
	\item[1-3] For every schematic mutex $\tup{p,q}$, where both $p$ and $q$ belong to $\in{\mathcal I}_{\xi,\Psi}$, one conditional effect is added to the ${\tt \small (insert\_pre)_{\xi,p}}$ actions to prevent the insertion of two preconditions that are {\em schematic mutex}. Likewise, two conditional effects are added to the ${\tt \small (insert\_eff)_{\xi,p}}$ actions, one to prevent the insertion of two positive effects that are {\em schematic mutex} and another one to prevent two mutex negative effects.
	\item[4-5] For every schematic mutex $\tup{p,q}$, where both $p$ and $q$ belong to $\in{\mathcal I}_{\xi,\Psi}$, two conditional effects are added to the ${\tt \small (apply)_{\xi,\omega}}$ actions to prevent positive effects that are inconsistent with an input observation (in ${\tt \small (apply)_{\xi,\omega}}$ actions the variables in $pars(\xi)$ are bounded to the objects in $\omega$ that appear in the same position).
\end{enumerate}

@@

The goals of the planning task $P'$ generated by the original BLS compilation are extended with the $\neg invalid$ literal to validate that only states consistent with the state constraints defined in $\Phi$ are traversed by solution plans. Remarkably, the $\neg invalid$ literal allows us also to define ${\tt \small (apply)_{\xi,\omega}}$ actions more compactly than in the original compilation. Disjunctions are no longer required to code the possible preconditions of an action schema since they can now be encoded with conditional effects of the type $\{pre\_\xi\_p\wedge \neg p(\omega)\}\rhd\{invalid\}$.

\subsection{Compilation properties}
\begin{mylemma}
Soundness. Any classical plan $\pi_{\Lambda}$ that solves $P_{\Lambda}$ produces a model $\mathcal{M'}$ that solves the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task.
\end{mylemma}

\begin{proof}[Proof]
\begin{small}
According to the $P_{\Lambda}$ compilation, once a given precondition or effect is inserted into the domain model $\mathcal{M}$ it cannot be removed back. In addition, once an action model is applied it cannot be modified. In the compiled planning problem $P_{\Lambda}$, only ${\tt \small (apply)_{\xi,\omega}}$ actions can update the value of the state fluents $F$. This means that a state consistent with an observation $s_n^o$ can only be achieved executing an applicable sequence of ${\tt \small (apply)_{\xi,\omega}}$ actions that, starting in the corresponding initial state $s_0^o$, validates that every generated intermediate state $s_i$, s.t. $0\leq i\leq n$, is consistent with the input state observations and {\em state-invariants}. This is exactly the definition of the solution condition for model $\mathcal{M}'$ to solve the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task.
\end{small}
\end{proof}

\begin{mylemma}
Completeness. Any model $\mathcal{M}'$ that solves the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task can be computed with a classical plan $\pi_{\Lambda}$ that solves $P_{\Lambda}$.
\end{mylemma}

\begin{proof}[Proof]
\begin{small}
By definition ${\mathcal I}_{\xi,\Psi}$ fully captures the set of elements that can appear in an action model $\xi$ using predicates $\Psi$. In addition the $P_{\Lambda}$ compilation does not discard any possible domain model $\mathcal{M}'$ definable within ${\mathcal I}_{\xi,\Psi}$ that satisfies the mutexes in $\Phi$. This means that, for every model $\mathcal{M}'$ that solves the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$, we can build a plan $\pi_{\Lambda}$ that solves $P_{\Lambda}$ by selecting the appropriate ${\tt \small (insert\_pre)_{\xi,e}}$ and ${\tt \small (insert\_eff)_{\xi,e}}$ actions for {\em programming} the precondition and effects of the corresponding action models in $\mathcal{M}'$ and then, selecting the corresponding ${\tt \small (apply)_{\xi,\omega}}$ actions that transform the initial state observation $s_0^o$ into the final state observation $s_m^o$.
\end{small}
\end{proof}

The size of the classical planning task $P_{\Lambda}$ output by our compilation depends on the arity of the given {\em predicates} $\Psi$, that shape the propositional state variables $F$, and the number of parameters of the action models, $|pars(\xi)|$. The larger these arities, the larger $|{\mathcal I}_{\xi,\Psi}|$. The size of the ${\mathcal I}_{\xi,\Psi}$ set is the term that dominates the compilation size because it defines the $pre\_\xi\_e/eff\_\xi\_e$ fluents, the corresponding set of ${\tt \small insert}$ actions, and the number of conditional effects in the ${\tt \small (apply)_{\xi,\omega}}$ actions. Note that {\em typing} can be used straightforward to constrain the FOL interpretations of $\Psi$ over the parameters $pars(\xi)$ which significantly reduces $|{\mathcal I}_{\xi,\Psi}|$ and hence, the size of the classical planning task output by the compilation.

Classical planners tend to preffer shorter solution plans, so our compilation may introduce a bias to $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning tasks preferring solutions that are referred to action models with a shorter number of {\em preconditions}/{\em effects}. In more detail, all $\{pre\_\xi\_e, eff\_\xi\_e\}_{\forall e\in{\mathcal I}_{\xi,\Psi}}$ fluents are false at the initial state of our $P_\Lambda$ compilation so classical planners tend to solve $P_\Lambda$ with plans that require a smaller number of {\em insert} actions.

This bias could be eliminated defining a cost function for the actions in $P_\Lambda$ (e.g. ${\tt \small insert}$ actions have {\em zero cost} while ${\tt \small (apply)_{\xi,\omega}}$ actions have a {\em positive constant cost}). In practice we use a different approach to disregard the cost of {\em insert} actions since classical planners are not proficient at optimizing {\em plan cost} when there are zero-cost actions. Instead, our approach is to use a SAT-based planner~\cite{rintanen2014madagascar} that can apply all actions for inserting preconditions in a single planning step (these actions do not interact). Further, the actions for inserting action effects are also applied in another single planning step. The plan horizon for programming any action model is then always bound to 2, which significantly reduces the planning horizon. The SAT-based planning approach is also convenient because its ability to deal with classical planning problems populated with dead-ends and because symmetries in the insertion of preconditions/effects into an action model do not affect the planning performance.

@@

\subsection{Observability versus Knowledge}
In our first experiment we seek to answer the question of whether knowledge can substitute observations. To that end, we evaluate the following 4 settings:

\begin{itemize}
	\item \textbf{Neither observability nor knowledge:} This is the baseline setting where the input sample is reduced to the minimum and only the initial and final states are known ($\sigma = 0$).
	\item \textbf{Only knowledge:} Here we add domain-specific knowledge encoded as schematic mutexes to the baseline scenario.
	\item \textbf{Only observability:} In this one, instead of knowledge we use a more complete input example where part of the intermediate states is known ($\sigma = 0.2$).
	\item \textbf{Both observability and knowledge:} In the last setting we use both more complete input examples and schematic mutexes.
\end{itemize}

@@

Table \ref{tab:observability_vs_knowledge} compiles the average precision (P) and recall (R) for each domain in the different settings tested. The table also reports the number of schematic mutexes ($|\Phi|$) used for each domain. Comparing the settings wtih only knowledge and only observability, it is clear that having more complete training examples is preferable. In fact, the improvement of using domain knowledge is marginal with respect to the baseline case. This is not the case for the last setting, where the use of domain knowledge shows a significant improvement in the quality of the learned models when compared to the setting with only observability.

@@

\section{Related work}
\label{sec:related}
In {\em Inductive Logic Programming} it is common to make the hypothesis be consistent with the {\em background knowledge}, that is some form {\em deductive knowledge} apart from the examples~\cite{muggleton1994inductive}.

{\em State-invariants} have also been previously used to improve the automatic construction of HTN planning model~\cite{lotinac2016constructing}.

@@

\section{Conclusions}
\label{sec:conclusions}
In some contexts it is however reasonable to assume that the action model is not learned from scratch, e.g. because some parts of the action model are known~\cite{ZhuoNK13,sreedharan2018handling,pereira2018heuristic}. Our compilation is also flexible to this particular learning scenario. The known preconditions and effects are encoded setting the corresponding fluents $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\xi,\Psi}}$ to true in the initial state. Further, the corresponding insert actions, $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertEff_{p,\xi}}$, become unnecessary and are removed from $A_{\Lambda}$, making the classical planning task $P_{\Lambda}$ easier to be solved. For example, suppose that the preconditions of the {\em blocksworld} action schema {\tt stack} are known, then the initial state is extended with literals, {\small\tt(pre\_holding\_v1\_stack)} and {\small\tt(pre\_clear\_v2\_stack)} and the associated actions $\mathsf{insertPre_{holding_v1,stack}}$ and $\mathsf{insertPre_{clear_v2,stack}}$ can be safely removed from the $A_{\Lambda}$ action set without altering the {\em soundness} and {\em completeness} of the $P_{\Lambda}$ compilation.

>>>>

\begin{enumerate}
	\item[1-3] For every schematic mutex $\tup{p,q}$, where both $p$ and $q$ belong to $\in{\mathcal I}_{\xi,\Psi}$, one conditional effect is added to the ${\tt \small (insert\_pre)_{\xi,p}}$ actions to prevent the insertion of two preconditions that are schematic mutex. Likewise, two conditional effects are added to the ${\tt \small (insert\_eff)_{\xi,p}}$ actions, one to prevent the insertion of two positive effects that are schematic mutex and another one to prevent two mutex negative effects.
	\item[4-5] For every schematic mutex $\tup{p,q}$, where both $p$ and $q$ belong to $\in{\mathcal I}_{\xi,\Psi}$, two conditional effects are added to the ${\tt \small (apply)_{\xi,\omega}}$ actions to prevent positive effects that are inconsistent with an input observation (in ${\tt \small (apply)_{\xi,\omega}}$ actions the variables in $pars(\xi)$ are bounded to the objects in $\omega$ that appear in the same position).
\end{enumerate}

@@

%The goals of the planning task $P'$ generated by the original BLS compilation are extended with the $\neg invalid$ literal to validate that only states consistent with the state constraints defined in $\Phi$ are traversed by solution plans. Remarkably, the $\neg invalid$ literal allows us also to define ${\tt \small (apply)_{\xi,\omega}}$ actions more compactly than in the original compilation. Disjunctions are no longer required to code the possible preconditions of an action schema since they can now be encoded with conditional effects of the type $\{pre\_\xi\_p\wedge \neg p(\omega)\}\rhd\{invalid\}$.

\subsection{Compilation properties}
\begin{mylemma}
Soundness. Any classical plan $\pi$ that solves $P'$ (planning task that results from the compilation) produces a model $\mathcal{M'}$ that solves the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task.
\end{mylemma}

\begin{proof}[Proof]
\begin{small}
According to the $P'$ compilation, once a given precondition or effect is inserted into the domain model $\mathcal{M}$ it cannot be undone. In addition, once an action model is applied it cannot be modified. In the compiled planning task $P'$, only ${\tt \small (apply)_{\xi,\omega}}$ actions can update the value of the state fluents $F$. This means that a state consistent with an observation $s_n^o$ can only be achieved executing an applicable sequence of ${\tt \small (apply)_{\xi,\omega}}$ actions that, starting in the corresponding initial state $s_0^o$, validates that every generated intermediate state $s_i$, s.t. $0\leq i\leq n$, is consistent with the input state observations and {\em state-invariants}. This is exactly the definition of the solution condition for model $\mathcal{M}'$ to solve the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task.
\end{small}
\end{proof}

\begin{mylemma}
Completeness. Any model $\mathcal{M}'$ that solves the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task can be computed with a classical plan $\pi$ that solves $P'$.
\end{mylemma}

\begin{proof}[Proof]
\begin{small}
By definition ${\mathcal I}_{\xi,\Psi}$ fully captures the set of elements that can appear in an action model $\xi$ using predicates $\Psi$. In addition the $P'$ compilation does not discard any model $\mathcal{M}'$ definable within ${\mathcal I}_{\xi,\Psi}$ that satisfies the mutexes in $\Phi$. This means that, for every model $\mathcal{M}'$ that solves the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$, we can build a plan $\pi$ that solves $P'$ by selecting the appropriate ${\tt \small (insert\_pre)_{\xi,e}}$ and ${\tt \small (insert\_eff)_{\xi,e}}$ actions for programming the precondition and effects of the corresponding action models in $\mathcal{M}'$ and then, selecting the corresponding ${\tt \small (apply)_{\xi,\omega}}$ actions that transform the initial state observation $s_0^o$ into the final state observation $s_m^o$.
\end{small}
\end{proof}

The size of $P'$ depends on the arity of the predicates in $\Psi$, that shape variables $F$, and the number of parameters of the action models, $|pars(\xi)|$. The larger these arities, the larger $|{\mathcal I}_{\xi,\Psi}|$. The size of ${\mathcal I}_{\xi,\Psi}$ is the most dominant factor of the compilation because it defines the $pre\_\xi\_e/eff\_\xi\_e$ fluents, the corresponding set of ${\tt \small insert}$ actions, and the number of conditional effects in the ${\tt \small (apply)_{\xi,\omega}}$ actions. Note that {\em typing} can be used straightforward to constrain the FOL interpretations of $\Psi$ over the parameters $pars(\xi)$, which will significantly reduce $|{\mathcal I}_{\xi,\Psi}|$ and hence the size of $P'$ output by the compilation.

Classical planners tend to prefer shorter solution plans, so our compilation may introduce a bias to $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning tasks preferring solutions that are referred to action models with a shorter number of preconditions/effects. In more detail, all $\{pre\_\xi\_e, eff\_\xi\_e\}_{\forall e\in{\mathcal I}_{\xi,\Psi}}$ fluents are false at the initial state of our $P'$ compilation so classical planners tend to solve $P'$ with plans that require a smaller number of ${\tt{\small insert}}$ actions.

This bias can be eliminated defining a cost function for the actions in $P'$ (e.g. ${\tt \small insert}$ actions have {\em zero cost} while ${\tt \small (apply)_{\xi,\omega}}$ actions have a {\em positive constant cost}). In practice we use a different approach to disregard the cost of ${\tt \small insert}$ actions since classical planners are not proficient at optimizing plan cost with zero-cost actions. Instead, our approach is to use a SAT-based planner~\cite{rintanen2014madagascar} that can apply all actions for inserting preconditions in a single planning step (these actions do not interact). Further, the actions for inserting action effects are also applied in another single planning step. The plan horizon for programming any action model is then always bounded to 2. The SAT-based planning approach is also convenient for its ability to deal with planning problems populated with dead-ends and because symmetries in the insertion of preconditions/effects into an action model do not affect the planning performance.

@@

\subsection{Observability versus Knowledge}
In our first experiment we seek to answer the question of whether domain knowledge can substitute observations. To this end, we evaluate the following 4 settings:

\begin{itemize}
	\item \textbf{Neither observability nor knowledge:} This is the baseline setting where the input sample is reduced to the minimum and only the initial and final states are known ($\sigma = 0$).
	\item \textbf{Only knowledge:} Here we add domain knowledge encoded as schematic mutexes to the baseline scenario.
	\item \textbf{Only observability:} In this one, we use a more complete observation where part of the intermediate states is known ($\sigma = 0.2$).
	\item \textbf{Both observability and knowledge:} In the last setting we use we use both more complete input examples and schematic mutexes.
\end{itemize}

@@

Table \ref{tab:observability_vs_knowledge} compiles the average precision (P) and recall (R) for each domain in the different settings tested. The table also reports the number of schematic mutexes ($|\Phi|$) used for each domain. Comparing the settings wtih only knowledge and only observability (second and third columns), it is clear that having more complete training examples is preferable. In fact, the improvement of using domain knowledge is marginal with respect to the baseline case. This is not the case for the last setting, where the use of domain knowledge shows a significant improvement in the quality of the learned models when compared to the setting with only observability.

@@

\section{Conclusions}
\label{sec:conclusions}

@@

In some contexts it is however reasonable to assume that the action model is not learned from scratch, e.g. because some parts of the action model are known~\cite{ZhuoNK13,sreedharan2018handling,pereira2018heuristic}. Our compilation is also flexible to this particular learning scenario. The known preconditions and effects are encoded setting the corresponding fluents $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\xi,\Psi}}$ to true in the initial state. Further, the corresponding insert actions, $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertEff_{p,\xi}}$, become unnecessary and are removed from $A_{\Lambda}$, making the classical planning task $P_{\Lambda}$ easier to be solved. For example, suppose that the preconditions of the {\em blocksworld} action schema {\tt stack} are known, then the initial state is extended with literals, {\small\tt(pre\_holding\_v1\_stack)} and {\small\tt(pre\_clear\_v2\_stack)} and the associated actions $\mathsf{insertPre_{holding_v1,stack}}$ and $\mathsf{insertPre_{clear_v2,stack}}$ can be safely removed from the $A_{\Lambda}$ action set without altering the {\em soundness} and {\em completeness} of the $P_{\Lambda}$ compilation.
